Phylogenetic inference of contact network parameters with approximate Bayesian computation

### 2016-06-27 doc/defence/presentation.tex
Paper and defence presentation updates

\documentclass{beamer}
\usepackage{tikz}
\usepackage{gillius}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage{multirow}
\usepackage{qrcode}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}

@@

    \uncover<9->{
      \node (n5) at (-0.86, 0.5) [circle, fill=black] { };
      \draw (n5) -- (n2);
      \draw (n5) -- (n1);
    }

    \uncover<10->{
      \node (n6) at (-1, 0) [circle, fill=black] { };
      \draw (n6) -- (n3);
      \draw (n6) -- (n4);
    }

    \uncover<11->{
      \node (n7) at (-0.86, -0.5) [circle, fill=black] { };
      \node (n8) at (-0.5, -0.86) [circle, fill=black] { };
      \node (n9) at (0, -1) [circle, fill=black] { };
      \node (n10) at (0.5, -0.86) [circle, fill=black] { };
      \node (n11) at (0.86, -0.5) [circle, fill=black] { };

      \draw (n7) -- (n2);
      \draw (n7) -- (n4);
      \draw (n8) -- (n2);
      \draw (n8) -- (n1);
      \draw (n9) -- (n2);
      \draw (n9) -- (n7);
      \draw (n10) -- (n3);
      \draw (n10) -- (n8);
      \draw (n11) -- (n2);
      \draw (n11) -- (n7);
    }
  \end{tikzpicture}

  \column{0.4\textwidth}
  \uncover<2->{
    $m$ = number of new edges per vertex \\
  }

  \uncover<3->{
    $\alpha$ = preferential attachment power \\
  }

  \uncover<11->{
    $N$ = number of nodes
  }
  \end{columns}
\end{frame}

\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{minipage}[p][\textheight][t]{\textwidth}
        \centering
        \only<1>{\includegraphics[page=1, width=\textwidth]{pa.pdf}}
        \only<2>{\includegraphics[page=2, width=\textwidth]{pa.pdf}}
        \only<3>{\includegraphics[page=3, width=\textwidth]{pa.pdf}}
        \only<4>{\includegraphics[page=4, width=\textwidth]{pa.pdf}}
        \only<5>{\includegraphics[page=5, width=\textwidth]{pa.pdf}}
        \only<6>{\includegraphics[page=6, width=\textwidth]{pa.pdf}}
      \end{minipage}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{minipage}[t][\textheight][t]{\textwidth}
      \begin{itemize}
        \uncover<1->{ \item Start with a small number of connected nodes. }
        \uncover<2->{ \item Attach new nodes with $m$ edges. }
        \uncover<3->{ \item Other endpoints of degree $d$ are chosen with probability $\propto d^{\alpha} + 1$. }
        \uncover<5->{ \item Continue until network has $N$ nodes. }
        \uncover<6->{ \item Also consider the prevalence $I$. }
      \end{itemize}
      \end{minipage}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Phylogenetic methods may help fit network models}
\end{frame}

@@

\begin{frame}{Transmission trees shape viral phylogenies}
\end{frame}

\begin{frame}{Can we fit network models with viral phylogenies?}
\end{frame}

%\begin{frame}{Preferential attachment generates scale-free networks}
%  \includegraphics[width=\textwidth]{pa-example}
%  \begin{itemize}
%    \item Sexual networks are often scale-free: degree distributions appear
%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
%      \pause
%    \item Preferential attachment models generate scale-free networks.
%  \end{itemize}
%\end{frame}

\begin{frame}{Do network parameters measurably affect tree shape?}
  \only<1>{
    \includegraphics[height=1in, trim=0 2in 4in 0, clip]{kernel-idea}
  }
  \only<2>{
    \includegraphics[height=1in, trim=0 2in 2in 0, clip]{kernel-idea}
  }
  \only<3->{
    \includegraphics[height=1in, trim=0 2in 0 0, clip]{kernel-idea}
  }
  \begin{itemize}
    \setlength{\itemsep}{0pt}
    \item Generate networks under different parameter values (number of nodes
      $N$, number of edges per vertex $m$, pereferential attachment power
      $\alpha$).
    \uncover<2->{
    \item Simulate epidemic over each network until $I$ nodes are infected.
    }
    \uncover<3->{
    \item Randomly subsample to form transmission trees.
    }
    \uncover<4->{
    \item Compare trees pairwise using tree kernel\footfullcite{poon2013mapping}.
    }
  \end{itemize}
\end{frame}

\begin{frame}{Preferential attachment power $\alpha$ affects tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 2.7in 3.3in 0, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-alpha-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Prevalence $I$ affects tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 2.7in 0in 0, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-I-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Number of edges per vertex $m$ does not affect tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 0.3in 3.3in 2.2in, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-m-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Number nodes $N$ modestly affects tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 0.3in 0 2.2in, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-N-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Simulated trees can be used to fit a network model}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
    \uncover<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
    \uncover<5->{\vspace{0.5cm}\centerline{\textbf{kernel approximate Bayesian computation (kernel-ABC)}}}
  \end{minipage}
\end{frame}

\begin{frame}{Example simulation shows $\alpha$ and $I$ can be reconstructed}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \begin{center}
    preferential attachment power \hfill prevalence
    \only<1>{
      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
    }
    \only<2>{
      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}

@@

\begin{frame}{Maximum \textit{a posteriori} estimates are accurate for $\alpha$ and $I$}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \vspace{-0.5cm}
    preferential attachment power \hfill \uncover<2>{edges per vertex}

    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m2.pdf}}

\begin{frame}{Real world HIV datasets exhibit network heterogeneity}
  \uncover<2->{
    \centerline{\includegraphics[height=0.8\textheight]{realdata-hpd-bc}}
  }

%  \scriptsize
%    \begin{tabular}{ccccc}
%      Reference & Sequences ($n$) & Location & Risk group & Gene \\
%      \hline
%      \color{magenta} \textcite{novitsky2013phylogenetic} &
%        \color{magenta} \multirow{2}{*}{180} &
%        \color{magenta} \multirow{2}{*}{Mochudi, Botswana} & 
%        \color{magenta} \multirow{2}{*}{HET} &
%        \color{magenta} \multirow{2}{*}{\textit{env}} \\ 
%      \color{magenta} \textcite{novitsky2014impact} \\
%      \color{blue} \textcite{cuevas2009hiv} & 
%        \color{blue} 287 & 
%        \color{blue} Basque Country, Spain & 
%        \color{blue} mixed & 
%        \color{blue} \textit{pol} \\
%      \color{cyan} \textcite{li2015hiv} & 
%        \color{cyan} 280 & 
%        \color{cyan} Shanghai, China & 
%        \color{cyan} MSM & 
%        \color{cyan} \textit{pol} \\
%      \color{green} unpublished & 
%        \color{green} 399 & 
%        \color{green} British Columbia, Canada & 
%        \color{green} IDU & 
%        \color{green} \textit{pol} \\
%      \color{orange} \textcite{wang2015targeting} & 
%        \color{orange} 173 & 
%        \color{orange} Beijing, China & 
%        \color{orange} MSM & 
%        \color{orange} \textit{pol} \\
%      \color{red} \textcite{niculescu2015recent} & 
%        \color{red} 136 & 
%        \color{red} Romaina & 
%        \color{red} IDU & 
%        \color{red} \textit{pol} \\
%      \hline
%    \end{tabular}
%  \normalsize
\end{frame}

\begin{frame}{Conclusions}
  \begin{itemize}
    \item Kernel-ABC is the first phylodynamic method to fit contact network
      models to phylogenetic data.
      \pause
    \item The preferential attachment power of the Barab\'asi-Albert network
      model, which is challenging to estimate by traditional epidemiological
      methods, can be estimated with kernel-ABC.
      \pause
    \item The networks underlying real epidemics are heterogeneous,
      underscoring the importance of considering network structure in
      phylodynamic analyses.
      \pause
  \end{itemize}
  \centerline{\qrcode{github.com/rmcclosk/netabc} github.com/rmcclosk/netabc}
\end{frame}

@@

\begin{frame}{Classifiers for network parameters}
  \vspace{-1cm}
  \begin{itemize}
    \item kSVR using tree kernel
    \item SVR using normalized
      lineages-through-time\footfullcite{janzen2015approximate}
    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
  \end{itemize}

  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
\end{frame}

\begin{frame}{Sequential Monte Carlo for ABC}
  \begin{tikzpicture}
    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
  \end{tikzpicture}
\end{frame}

>>>>

\documentclass{beamer}
\usepackage{tikz}
\usepackage{gillius}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage{multirow}
\usepackage{qrcode}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}

@@

    \uncover<9->{
      \node (n6) at (-0.86, 0.5) [circle, fill=black] { };
      \draw (n6) -- (n2);
      \draw (n6) -- (n1);
    }

    \uncover<10->{
      \node (n7) at (-1, 0) [circle, fill=black] { };
      \draw (n7) -- (n3);
      \draw (n7) -- (n4);
    }

    \uncover<11->{
      \node (n8) at (-0.86, -0.5) [circle, fill=black] { };
      \node (n9) at (-0.5, -0.86) [circle, fill=black] { };
      \node (n10) at (0, -1) [circle, fill=black] { };
      \node (n11) at (0.5, -0.86) [circle, fill=black] { };
      \node (n12) at (0.86, -0.5) [circle, fill=black] { };

      \draw (n8) -- (n2);
      \draw (n8) -- (n4);
      \draw (n9) -- (n2);
      \draw (n9) -- (n1);
      \draw (n10) -- (n2);
      \draw (n10) -- (n8);
      \draw (n11) -- (n3);
      \draw (n11) -- (n9);
      \draw (n12) -- (n2);
      \draw (n12) -- (n8);
    }

    \uncover<12->{
      \node (n1) at (1, 0) [circle, fill=red] { };
      \node (n2) at (0.86, 0.5) [circle, fill=red] { };
      \node (n5) at (-0.86, 0.5) [circle, fill=red] { };
      \node (n7) at (-0.86, -0.5) [circle, fill=red] { };
      \node (n8) at (-0.5, -0.86) [circle, fill=red] { };

      \draw [ultra thick, red] (n2) -- (n7);
      \draw [ultra thick, red] (n2) -- (n1);
      \draw [ultra thick, red] (n8) -- (n1);
      \draw [ultra thick, red] (n5) -- (n2);
    }

    \only<11>{\node [below=of n10] {$N = 12$};}
    \uncover<12->{\node [below=of n10] {$I = 5$};}
  \end{tikzpicture}

  \column{0.4\textwidth}
  \uncover<2->{
    $m$ = number of new edges per vertex \\
    \hfill\\
  }

  \uncover<3->{
    $\alpha$ = preferential attachment power \\
    \hfill\\
  }

  \uncover<11->{
    $N$ = number of nodes \\
    \hfill\\
  }

  \uncover<12->{
    $I$ = number of infected nodes (prevalence) 
  }
  \end{columns}
\end{frame}

@@

\begin{frame}{Phylogenetic methods may help fit network models}
  \begin{itemize}
    \item network models are difficult to fit with conventional epidemiological
      methods (surveys etc.)
    \item sequence data may be easier to collect and free from some biases
      (misreporting etc.)
    \item instead of asking the people, ask the virus! (some cartoon)
  \end{itemize}
\end{frame}

@@

\begin{frame}{Transmission trees shape viral phylogenies}
  some demo of viruses evolving and being transmitted at the same time
\end{frame}

\begin{frame}{Can we fit network models with viral phylogenies?}
  \begin{center}
  network $\to$ transmission tree $\to$ viral phylogeny \\
  \hfill\\
  network $\stackrel{\alert{?}}{\leftarrow}$ transmission tree $\leftarrow$ viral phylogeny \\
  \end{center}
\end{frame}

\begin{frame}{Are network parameters identifiable from tree shape?}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{kernel-idea.pdf}}
    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.7in 0 0, clip]{kernel-idea.pdf}}
    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 0.8in 0 0, clip]{kernel-idea.pdf}}
    \only<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{kernel-idea.pdf}}
  \end{minipage}
\end{frame}

\begin{frame}{yes they do}
  \centerline{\includegraphics[height=0.8\textheight]{kernel-kpca}}
\end{frame}

\begin{frame}{Simulated trees can be used to fit network models}
  \centerline{\includegraphics[width=0.9\textwidth]{abc-smc}}

  \centerline{animate this?}
\end{frame}

\begin{frame}{Posterior mean estimates are more accurate for $\alpha$ and $I$}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \vspace{-0.5cm}
    preferential attachment power \hfill \uncover<2>{edges per vertex}

    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m3.pdf}}
    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m3.pdf}}

\begin{frame}{Example simulation\ldots}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \begin{center}
    preferential attachment power \hfill prevalence
    \only<1>{
      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
    }
    \only<2>{
      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}

@@

\begin{frame}{Example simulation continued, $N$ and $I$ confounded}
  %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
\end{frame}

\begin{frame}{Application to real world HIV datasets}
  \centerline{\input{\tablepath/realdata-bc-short}}
\end{frame}

\begin{frame}{HIV networks have sub-linear preferential attachment}
    Also, higher PA for IDU networks.

    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
\end{frame}

\begin{frame}{Conclusions}
  \begin{itemize}
    \item We developed a phylodynamic method to fit contact network models to
      phylogenetic data.
      \pause
    \item The preferential attachment power of the Barab\'asi-Albert network
      model, which is challenging to estimate by traditional epidemiological
      methods, can be estimated with ABC.
      \pause
    \item The networks underlying real epidemics are heterogeneous,
      underscoring the importance of considering network structure in
      phylodynamic analyses.
  \end{itemize}
\end{frame}

@@
